Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2022-03-01 unimodular-covers-CombTheory-submitted.tex
Add files via upload

\usepackage[colorinlistoftodos]{todonotes}

\newcommand{\giulia}[1]{\todo[size=\tiny,color=blue!30]{#1 \\ \hfill --- G.}}
\newcommand{\Giulia}[1]{\todo[size=\tiny,inline,color=blue!30]{#1 \\ \hfill --- G.}}

\newcommand{\paco}[1]{\todo[size=\tiny,color=green!30]{#1 \\ \hfill --- P.}}
\newcommand{\Paco}[1]{\todo[size=\tiny,inline,color=green!30]{#1 \\ \hfill --- P.}}

@@

\author[G.~Codenotti]{Giulia Codenotti}
\author[F.~Santos]{Francisco Santos}

\address[G.~Codenotti]
{
Institut f\"ur Mathematik, Freie Universit\"at Berlin, Germany
}
\email{giulia.codenotti@fu-berlin.de}

\address[F.~Santos]
{
Department of Mathematics, Statistics and Computer Science, University of Cantabria, Spain
}
\email{francisco.santos@unican.es}

\thanks{The authors were supported by the Einstein Foundation Berlin under grant EVF-2015-230. 
Work of F. Santos is also supported by grants PID2019-106188GB-I00 funded by MCIN/AEI/10.13039/501100011033 and by FPU19/04163 of the Spanish Government and by project CLaPPo (21.SI03.64658) of Universidad de Cantabria and Banco Santander}

@@

\title{Unimodular covers of 3-dimensional parallelepipeds and Cayley sums}

@@

%\tableofcontents

@@

When the normal fan of a polytope $Q$ refines that of another polytope $P$, as in the second conjecture, we say that $P$ \emph{is a weak Minkowski summand of $Q$}, since this is easily seen to be equivalent to the existence of a polytope $P'$ such that $P+P' = k Q$ for some dilation constant $k>0$.
%\paco{did not find this in the google books version of \cite{Grunbaum}; found it in Postnikov-Reiner-Williams, Tsuchiya, a survey by Rekha Thomas, and other places. Perhaps we can leave it without reference} 
This property has the following algebraic implication for the projective toric variety $X_Q$: $P$ is a weak Minkowski summand of $Q$ if and only if the Cartier divisor defined by $P$ on  $X_Q$ is \emph{numerically effective}, or ``nef'' (see~\cite[Cor.~6.2.15, Prop.~6.3.12]{CLS}, but observe that what we here call  ``weak Minkowski summand'' is simply called ``Minkowski summand'' there).

@@

This has the following consequence, also proved in \Cref{sec:cayley}.
Here a  \emph{prismatoid} is a polytope whose vertices all lie in two parallel facets. 
%A polytope has width $1$ if its vertices lie in two \emph{consecutive} parallel lattice hyperplanes. Observe that this is the same as being ($SL(\Z,d)$-equivalent to) a Cayley sum.

@@

%\paco{deleted former corollary}
%\Paco{
%I am a bit puzzled. After the next corollary we say that \cite[Cor.~4.2]{SantosZiegler} proves a case of this, but in fact it proves the whole of it, in a stronger form: it does not need width one, and it gives a "dissection" and not a mere cover (dissection= cover by simplices with disjoint interiors). 
%I propose to delete it:
%}
%{\color{red}
%\begin{corollary}
%%\label{coro:width1}
%Every integer dilation $kP$, $k\ge 2$, of a lattice $3$-polytope $P$ of width $1$ has a unimodular cover.
%\end{corollary}
%
%A special case of the latter are integer dilations of empty tetrahedra. That their dilations have
%unimodular covers is \cite[Cor.~4.2]{SantosZiegler} (and is also implicit in \cite{KantorSarkaria}).
%}

\paco{added ref to Gubeladze}
\giulia{changed it}
\paco{changed again, adding more references}

Let us mention that recent work of Gubeladze~\cite{Gubeladze} shows another class of 3-polytopes admitting unimodular covers: the convex hulls of all lattice points inside an ellipsoid; these had previously been shown to be IDP by Bruns, Gubeladze and Michalek~\cite{BGM}.
To date there are no known examples of IDP $3$-polytopes without a unimodular cover, although such polytopes exist in higher dimension~\cite{BG}.
%Let us also mention the following recent results of Gubeladze~\cite{Gubeladze}.
%In part (2), a polytope $P$ is called \emph{very ample} if the vertex cone $\cone(P-v)$ is contained in the translation $P-v$, for every vertex $v$ of $P$.

%\begin{theorem}[Gubeladze~\cite{Gubeladze}]
%Let $P$ be a lattice $3$-polytope. 
%\begin{enumerate}
%\item If $P=E\cap \Z^3$ for some (solid) ellipsoid $E$, then $P$ has a unimodular cover.
%\item If $P$ is \emph{very ample} then the boundary of $P$ has a unimodular cover. That is, there are unimodular simplices contained in $P$ whose union contains $\partial P$.
%\end{enumerate}
%\end{theorem}

%\paco{removed closing sentence}
%\medskip
%We believe that the $3$-polytopes in all these statements have unimodular triangulations, but this remains an open question.

\subsection*{Acknowledgements:} We would like to thank Akiyoshi Tsuchiya, Spencer Backman, and Johannes Hofscheier for posing these questions to us and Christian Haase for helpful discussions.

@@

Observe that $u+v=p_1+p_2=q_3+q_4$.
Now, this implies that the quadrilateral $\conv(p_1q_4p_2q_3)$ contains a fundamental domain for the lattice $\Z^2\times\{0\}$. Hence, its translate $\conv(q_2p_3q_1p_4)$ contains a fundamental domain for $\Z^2\times\{1\}$ and, in particular, it contains at least one lattice point other than $p_3$ and $p_4$. By central symmetry around its center $\left(\frac{a}2,-\frac{b}2,1\right)$, $\conv(q_2p_3q_1p_4)$ must contain lattice points in both triangles $\conv(q_2p_3p_4)\subset T_1$ and $\conv(q_1p_3p_4)\subset T_2$.
%

%\begin{figure}[htb]
%\includegraphics[scale=.28]{Whites}
%\caption{The parallelograms $\conv(p_1p_2q_3q_4)$ and $\conv(q_2q_1p_4p_3)$. The large red dots are lattice points.}
%\label{fig:corner}
%\end{figure}

@@

The translation of vector $\frac12 (p_1+p_2-p_3-p_4)$ sends $q_1,q_2,p_3,p_4$ to $p_2,p_1,q_4,q_3$ (in this order). By applying this to inequality \eqref{eq:third}, we obtain
\[
\alpha \le f(p_2) < f(q_i) < f(p_1) \le \beta, \quad\text{ for $i=3,4$},
\]
so that $q_3,q_4 \in B$.
%
This finishes the proof of the claim, and of the lemma.
%
%Let $f\in (\R^3)^*$ be any functional. Fix $i \in [4]$; observe that if $f(q_i) > f( p_j)$ for all $j\neq i$, then 
%\begin{gather*}
%f(q_i) \overset{(i)}{>} f(q_j) \overset{(ii)}{>} f(p_i) \\
%f(q_i) \overset{(iii)}{>} f(p_j) \overset{(iv)}{>} f(p_i) \\
%f(p_j)\overset{(v)}{>}f(q_k)
%\end{gather*}
%for all $j \neq i$, $k \notin \{j, i\}$. 
%
%Indeed, for any $\{i,j,k,l\}=[4]$ we have 
%
%\begin{align*}
%f(q_i) > f( p_j) \Leftrightarrow \frac12 f(p_l+&p_k -p_j-p_i) > 0  \Leftrightarrow f(q_j) > f(p_i)\\
%& \Updownarrow \\
% f(q_l) < f(p_k) \Leftrightarrow \frac12 f(-p_l-&p_k +p_j+p_i) < 0 \Leftrightarrow f(q_k) < f(p_l),
%\end{align*}
%which shows the equivalence of $(i)$, $(iii)$ and $(v)$; note that $(iii)$ is our assumption. Inequalities $(ii)$ and $(iv)$ follow by taking the sum of two inequalities of the form $\frac12 f(p_l+p_k -p_j-p_i) > 0$. 
%
%Using this observation, we can show that each of pairs of parallel facets of $P$ cuts out at most one of the $q_i$s:
%if $q_i$ lies outside of $P$, it must violate a facet-defining inequality $f(x) \leq a$.Thus $f(q_i) > a \geq f(p_j)$, and by our observation, it follows that all remaining $q_j$ satisfy the inequality $f(x) \leq a$. Since $P$ is a parallelogram, there is a facet with inequality $f(x) \geq b$; by what is said above, the minimum value of $f(x)$ is attained at $p_i$, which is in $P$. Thus $f(q_j) > f(p_i) \geq b$, that is, the inequality is satisfied by all the $q_j$.
%
%Since at most one of the $q_i$ does not satisfy the inequalities corresponding to a pairs of parallel facets of $P$, and there are only three pairs of facets, one of $q_i$s must satisfy all the inequalities and therefore be inside of $P$.
\end{proof}
%\giulia{here it may also be nice to add a picture}

@@

\Cref{lemma:3<4} guarantees that one of the corner tetrahedra $T_i$ of the parallelepiped $C(T)$ is contained in $P$. Without loss of generality, suppose $T_4 = \conv(p_1, p_2, p_3,q_4)$ is in $P$. By \Cref{lemma:corner}, we know that $T_4$ contains a lattice point other than the $p_i$s, which we denote by $u$. 
%Observe that from the proof of \Cref{lemma:corner} we know more: $u$ lies in the face $\conv(p_1, p_2, q_4)$.\giulia{we need this for the volume computation below. perhaps it should be made explicit in the statemente of the lemma...}
%
Then $S=\conv(T\cup \{u\})$ can be triangulated in two different ways: $S=T \cup T'_4$, where $T'_4 = \conv(p_1, p_2, p_3, u) \subseteq T_4$ and $S= S_1 \cup S_2 \cup S_3$, with
\[
S_1= \conv(p_2,p_3,p_4, u),
S_2=\conv(p_1,p_3,p_4, u),
S_3=\conv(p_1,p_2,p_4, u).
\]

Each of the tetrahedra $S_i$ has lattice volume strictly smaller than $T$ because, for each $i$, $p_i$ is the unique point of $C(T)$ maximizing the distance to the opposite facet $\conv(p_j,p_k,p_l)$ of $T$. Thus, $S_1$, $S_2$ and $S_3$ cover $T$ and have volume strictly smaller than $T$. The $S_i$ may not be empty, but we can triangulate them into empty tetrahedra, which by inductive hypothesis they can be covered unimodularly.
%
%Observe that  $T'_4$ and/or one of the $S_i$ can degenerate to a triangle, if $u$ lies in $\conv(p_1,p_2,p_3)$, but this does not affect the argument.
%
%: indeed, $\Vol(S_3)=\Vol(T'_4)$, since these tetrahedra have a common face $\conv(p_1, p_2, l)$ and the remaining vertex is at distance one from this face in both. Now $T'_4$ is contained in $T_4$, and we know that each corner tetrahedron $T_i$ has volume half that of $T$. The remaining two tetrahedra $S_1$ and $S_2$ must therefore have volume summing to that of $T$, and therefore each has volume strictly less than $T$.
\end{proof}

\begin{proof}[Proof of \Cref{thm:parallelepipeds}]
Arbitrarily triangulate the parallelepiped into empty lattice tetrahedra and apply \Cref{coro:coverpara} to these tetrahedra.
%Let $P \subseteq \R^3$ be a lattice parallelepiped. It can be triangulated into empty lattice tetrahedra, simply by taking a triangulation which uses all lattice points as vertices.  If all the tetrahedra in the triangulation are unimodular we are done. Otherwise, suppose $T$ is an empty tetrahedron in the triangulation which is not unimodular. \Cref{lemma:3<4} guarantees that one of the corner tetrahedra $T_i$ of the parallelepiped $C(T)$ circumscribed to $T$ is contained in $P$. Without loss of generality, suppose $T_4 = \conv(p_1, p_2, p_3,q_4)$ is in $P$. By \Cref{lemma:corner}, we know that $T_4$ contains a lattice point, which we denote by $u$. 
%
%
%This allows us to cover $T$ with lattice simplices of smaller volume recursively, until it is covered by unimodular ones.
\end{proof}

%\begin{remark}
%\label{rem:octahedra_and_tri-prisms}

%We finish with a question which, if answered in the positive, would imply \Cref{conj:Oda}\eqref{itm:smoothIDP} in dimension three.
Let us say that a lattice $3$-polytope $P$ \emph{has the circumscribed parallelepiped property} if it satisfies the conclusion of \Cref{lemma:3<4}:  ``for every empty tetrahedron $T$ contained in $P$ at least one of the four corner tetrahedra in $C(T)$ is 
contained in $P$''. 
If this holds then $P$ has a unimodular cover, since then the proofs of \Cref{coro:coverpara} and \Cref{thm:parallelepipeds} work for $P$.
In turn, our proof that parallelepipeds have the property (\Cref{lemma:3<4}) is based on the fact that they have only three (pairs of) normal vectors. 
In the following two examples we show a smooth $3$-polytope and two $3$-polytope with four normal vectors that do not have the property. The latter are not IDP:

@@

%The assumption that $P+Q$ is $d$-dimensional is equivalent to $\cayley (P,Q)$ being $(d+1)$-dimensional.

@@

\begin{lemma}
\label{lemma:cayley}
%\paco{$Q$ is 2-dim, but $P$ may not}
Let $Q$ be a two-dimensional lattice polytope and $P$ a weak Minkowski summand of it. 
Let $p=[p_1,p_2] \subset P$ and $q=[q_1,q_2]\subset Q$ be two primitive and non-parallel lattice segments, and let $\vecline p$ and $ \vecline q$ be the lines spanned by them.  If the parallelogram $p + q$ is not unimodular, then at least one of the regions
\[
((p_1, p_2) + \vecline q ) \cap P, 
\qquad \text{and} \qquad
((q_1, q_2) + \vecline p ) \cap Q
\]
contains a lattice point. See \Cref{fig:strips}.
%\[
%(\int(p) + \vecline q) \cap P
%\qquad \text{or} \qquad
%\int(q + \vecline p) \cap Q
%\]
%contains a lattice point.
\end{lemma}

@@

In turn, this means that the polytope $\conv(\tilde u, \tilde p_1, \tilde p_2, \tilde q_1, \tilde q_2) = \cayley(t,q)$
 has the following two triangulations:
\begin{gather*}
\mathcal T^+:= \left\{ \cayley(p,q), \cayley(t, \{q_i\}) \right\},
\\
\mathcal T^-:= \{ \cayley([p_1,u],q), \cayley([p_2,u],q), \cayley(t, \{q_j\}) \}.
\end{gather*}
The tetrahedra $\cayley(t, \{q_j\})$ and $\cayley(t, \{q_i\})$ are unimodular, which implies that $T=\cayley(p,q)$ has volume equal to the sum of the volumes of $\cayley([p_1,u],q)$ and $\cayley([p_2,u],q)$. In particular, we have covered $T$ by the three tetrahedra in $\mathcal T^-$, which are of smaller volume and hence have unimodular covers by inductive assumption.
%
%A unimodular triangulation of a simplex of type $(3,1)$ (resp $(1,3)$) is obtained simply by unimodularly triangulating the triangular face lying in $P$ (resp $Q$) and coning over the vertex in $Q$ (resp $P$). This leaves simplices of type $(2,2)$. In the mixed subdivision of $P+Q$ corresponding to $T$, these are mixed cells, that is parallelograms with one edge in $P$ and one in $Q$ \giulia{how imprecise can be without being confusing?}.
%
%Let $S=\{p_1, p_2\} + \{q_1, q_2\}$ be such a parallelogram. By \Cref{lemma:cayley} we know that there is a lattice point $p \in P$ such that either $p+q_1$ or $p+q_2$ lies in $S$, w.l.o.g. suppose it is $p+q_1$. Then we can cover $S$ by the triangle $\{p_1, p_2, p\} + q_1$ and parallelograms $\{p_1, p\}  + \{q_1, q_2\}$ and $\{p_2, p\} + \{q_1, q_2\}$. As above, the triangle can be unimodularly triangulated, while the volume of the two parallelograms sums up to the volume of $S$, and therefore each has volume strictly smaller than $S$. By iterating this procedure, we can cover $S$ by unimodular cells.
\end{proof}

@@

\begin{proof}[Proof of \Cref{coro:prismatoid}]
The polytope under study satisfies the hypotheses of \Cref{prop:prismatoid}: %in \Cref{coro:prismatoid}, 
the smoothness of the prismatoid implies that every edge of the form $[u\times \{0\}, v\times \{k\}]$ has lattice points in all slices. 
%In \Cref{coro:width1}, since $P$ has width one, $P\cong \cayley(Q_1, Q_2)$ for some $Q_1$ and $Q_2$. 
Hence,
\[
kP \cap(\R^2\times\{1\}) = (k-1)Q_1 + Q_2.
\qedhere
\] 
\end{proof}

@@

%\begin{claim}
%$b_2$ and $t_2$ are non-zero and have the same sign. W.l.o.g.~we assume they are positive.
%\end{claim}
%
%\begin{proof}
%If, say, $b_2 \le 0 \le t_2$. Then the intersections of $t$ and $b$ with the vertical line $\{x=-\beta\}$ through $q_1$ are less than distance one apart (because $t$ must go below every lattice point in the interior of $p_1p_2 + q_1q_2 - p_1$ and $b$ above every latttice point in the interior of $p_1p_2 + q_1q_2 - p_2$). This implies the width w.r.t to $f_q$, the functional constant on $q_1q_2$, of any segment with vertices in $Q \cap \{x\le  -\beta\}$ is less than one, so it is impossible to fit a translated copy of $l$ in it.
%\end{proof}

@@

%\begin{claim}
%The slopes of $b$ and $t$ must either be both strictly larger than the slope of $q$ or both be strictly smaller. W.l.o.g.~we assume they are larger.
%\end{claim}
%
%\begin{proof}
%If, say, $b$ has positive slope and $t$ negative. Then since $t$ intersects the vertical line $x=0$ spanned by $p$ below $p_2$, and $b$ intersects it above $p_1$, the slopes of $b$ and $t$ imply that

@@

%\Giulia{fix it to look more unimodular at the bottom}
%\begin{figure}[htb]
%  \centering
%\begin{tikzpicture}[mynode/.style={circle, inner sep=1.5pt, fill=red}, scale =
%  1]
%%coordinate axes
%   \tkzInit[xmax=3.5,ymax=3.5,xmin=-3.5,ymin=-3.5]
%%        \tkzLabelX[orig=false,label options={font=\tiny}]
%%        \tkzLabelY[orig=false,label options={font=\tiny}]
%%        \tkzGrid
%%        \tkzDrawX
%%        \tkzDrawY
%
%%edges b, t
%  \node[
%%label={[shift={(0,-0.67)}]$b_1$}, 
%mynode] (b1) at (-3,-2) {}; 
%  \node[
%%label={[shift={(0,-0.67)}]$b_2$}, 
%mynode] (b2) at (2, 0.2) {};  
%  \node[
%%label={[shift={(0,0)}]$t_1$}, 
%mynode] (t1) at (-2.7, 1) {};  
%  \node[
%%label={[shift={(0,0)}]$t_2$}, 
%mynode] (t2) at (1.6,0.7) {};  
%
%%shaded region where r can be  
%  \draw[fill=gray!80, nearly transparent] (2.93,0.61) -- (0,0.81)  -- (0,-0.68) -- cycle;
%
%
%%p
%  \node[
%%label={[shift={(0,-0.67)}]$p_1$}, 
%mynode] (p1) at (0,-1) {}; 
%  \node[
%%label={[shift={(0,0)}]$p_2$}, 
%mynode] (p2) at (0,1) {};  
%%  \node[{p}]  at (-0.3,0.95)
%
%
%%q
%  \node[
%%label={[shift={(0,-0.67)}]$q_1$}, 
%mynode] (q1) at (-1.7,-0.4) {}; 
%  \node[
%%label={[shift={(0,-0.67)}]$q_2$}, 
%mynode] (q2) at (0.3,-0.4) {};  
%
%%segments b, t, p, q
%  \draw[ name path = b, black, thick] (b1) node[below, xshift=1.7cm, yshift=0.7cm] {b} -- (b2); 
%  \draw[name path = t, black, thick] (t1) node[below, xshift=1.2cm, yshift=-0.1cm] {t} -- (t2);
%  \draw[name path = p, black, thick] (p1) node[left, xshift=0 cm, yshift=1.3cm] {p} -- (p2);
%  \draw[name path = q, black, thick] (q1) node[below, xshift=0.4cm, yshift=0cm] {q} -- (q2);
%
%
%%prolonged lines through b and t
% \tkzDrawLine[add=0.2 and 0.3, color=black, dashed](b1, b2);
% \tkzDrawLine[ add=0.3 and 0.4, color=black, dashed](t1, t2);
% 
%%functional f_q
%  \tkzDefLine[orthogonal=through p1](p1,p2)
%  \tkzDrawLine[add = 2 and 1,color=black, dashed](p1,tkzPointResult)
%  \node[label={\small $\{f_q(x)=-1\}$}, xshift=0.4cm, yshift=-0.2cm] (lab1) at (3.3, -1) {};
%
%
%  \tkzDefLine[orthogonal=through p2](p1,p2) 
%  \tkzDrawLine[add = 2 and 1,color=black, dashed](p2,tkzPointResult) 
%    \node[label={\small $\{f_q(x)=w-1\}$}, xshift=0.4cm, yshift=-0.2cm] (lab2) at (3.3, 1) {};
%
%\end{tikzpicture}
%\caption{Picture for Claim 4.2}
%\label{fig:claim2}
%\end{figure}

@@

%\begin{figure}
%\includegraphics[ width=.7\textwidth]{Claim43.jpg}
%\caption{Figure for Claim 4.3}
%\label{fig:claim3}
%\end{figure}
\begin{figure}[htb]
\scalebox{.75}{\input{claim3.pdf_t}}
\caption{Illustration of the proof of \Cref{claim:r}}
\label{fig:claim3}
\end{figure}

@@

%Starting here we assume index 2.
%
%\begin{claim}
%The slopes of $t$ and $b$ are strictly smaller than one and that of $r$ strictly greater than one.
%\end{claim}
%
%\begin{proof}
%Otherwise the points $(0,1/2)$, $(0,-1/2)$, $(1/2,0)$, respectively, would be in the corresponding band.
%\end{proof}
%
%We are now ready to show a contradiction. Let $u$ and $v$ be the end-points of the translated copy of $r$ in $Q$, with $\vec r=v-u$. Let $d$ be the line of slope $1$ through $u$, which is a lattice line. Let $d'$ be the next parallel lattice line above $d$. By Claim A.3.1, the intersection of $d'$ with the vertical line through $u$ is above $Q$. Since the slope of $t$ is less than one, the same happens for every point of $d'$ to the right of $u$. In particular, $v$ must be below $d'$ and hence it must be on or below $d$. This contradicts the fact that the slope of $r$ is strictly greater than one. QED

@@

\begin{claim}
\label{claim:g}
$g(t_1) > g(t_2)$, $g(b_1) > g(b_2)$, and $g(r_1) < g(r_2)$.
%The slopes of $t$ and $b$ are strictly smaller than the slope of  $[p_1, q_2]$ and that of $r=[r_1, r_2]$ (and $l=[l_1, l_2]$) is strictly greater than it.
\end{claim}

@@

%
%\appendix
%
%\section{Old stuff}
%
%
%\begin{proof}
%%Suppose by contradiction that there is no lattice point as described in the lemma. Then there cannot be a vertex of $P$ in the corresponding strip, since $P$ is a lattice polygon. The same holds for $Q$ and its corresponding strip. Thus $P$ has two edges which each have one vertex on each side of the strip $(p_1p_2 + \langle \overrightarrow{q_1q_2}\rangle)$; we will call these edges $e_1$ and $e_2$. The same holds for $Q$ and the strip $(q_1q_2 + \langle \overrightarrow{p_1p_2}\rangle)$, and we call the edges be $f_1, f_2$. 
%
%Let us fix coordinates in which $p_1p_2$ and $q_1q_2$ are a vertical and horizontal segment of length one intersecting at the origin. That is, there are numbers $\alpha, \beta \in (0,1)$ such that
%\[
%p_1=(0,-\alpha), \quad
%p_2=(0,1-\alpha), \quad
%q_1=(-\beta,0), \quad
%q_2=(1-\beta,0).
%\] 
%The ambient lattice is a certain lattice containing $p_1,p_2,q_1,q_2$ (the case of index two is when $\alpha=\beta=1/2$ and the lattice is spanned by these four points).
%
%Suppose by contradiction that there is no lattice point as described in the lemma. Then there cannot be a vertex of $P$ in the corresponding strip, since $P$ is a lattice polygon. The same holds for $Q$ and its corresponding strip. Thus $P$ has two primitive edges which each have one vertex on each side of the strip $p_1p_2 + \langle \overrightarrow{q_1q_2}\rangle$; we will call these edges $\ell$ and $r$ (with $\ell$ to the left and $r$ to the right). The same holds for $Q$ and the strip $q_1q_2 + \langle \overrightarrow{p_1p_2}\rangle$, and we call the edges  $b, t$ (with $b$ below and $t$ above). To make things more explicit, we assume the following coordinates for the edge-vectors:
%\[
%\vec l = (l_1,l_2), \quad
%\vec r= (r_1,r_2), \quad
%\vec b= (b_1,b_2), \quad
%\vec t= (t_1,t_2).
%\]
%Without loss of generality we assume $l_2,r_2,b_1,t_1 >0$. (They are non-zero and we can arbitrarily choose a sign for each vector).
%
%
%
%
%
%
%Since the normal fan of $Q$ refines that of $P$, $Q$ will contain edges $e_1'$ and $e_2'$ parallel to $e_1$ and $e_2$ respectively. These edges of $Q$ are all distinct and appear alternating in $Q$, that is, they appear in the order $f_1, e_1', f_2, e_2'$. Indeed, consider the quadrilateral $S$ with vertices $q_1, p_1+l, q_2, p_2+l$, where $l$ is a lattice point chosen so that the segments $q_1q_2$ and $p_1+l,p_2+l$ intersect in their interior; we can further require that $\conv( q_1, p_1+l, q_2)$ is unimodular \giulia{not sure if we use it}. $S$ lies inside the $Q$-strip, while its translate $S-l$ lies inside the $P$-strip. Thus in order to have no lattice points in the strips, we must have that the edge $f_i$ separates the vertex $p_i +l$ of $S$ from the remaining three, and  the same for the edge $e_i +l$ and vertex $q_i$ of $S$. This is equivalent to saying that the normals to the edges $f_1, e_1, f_2, e_2$ lie in different open cones of the normal fan of $S$, in this cyclic order. 
%
%- both $f_1$ or $f_2$ must meet the line through $q_1$ and $q_2$, else they would not cut out the lattice point they need to. same goes for $e_1$ and $e_2$ and the line through $p_1$ and $p_2$.
%
%claim: the segments $e_1 + l$ and $e_2 + l$ respectively intersect both $f_1$ and $f_2$. 
%
%claim: if we tilt one of the $f_i$s so that they are parallel, there is no room to fit $e_i$, since it intersects both. 
%
%
%
%
%
%%Consider the segments on the lattice lines parallel to $p_1 p_2$ cut out by the edges $f_1$ and $f_2$ \giulia{this should be formulated more comprehensibly}. Since these segments do not contain lattice points, they must have lattice length stricly less than $1$ (or the width wrt $q$ is less than $k$, the index of $p,q$), which means that the smallest width in direction of $p_1 p_2$ between a "consecutive" vertices of $f_1$ and of $f_2$ must also be strictly less than $k$. 
%
%%Further, we observe that both vertices of $f_1$ (or $f_2$) cannot lie strictly above the line through $q_1$ and $q_2$, else the strip would contain a lattice point. 
%\end{proof}
%
%
%
%\begin{figure}[htb]
%  \centering
%\begin{tikzpicture}[mynode/.style={circle, inner sep=1.5pt, fill=red}, scale =
%  .7]
%%coordinate axes
%   \tkzInit[xmax=3.5,ymax=3.5,xmin=-3.5,ymin=-3.5]

>>>>

@@

%\author[G.~Codenotti]{Giulia Codenotti}
%\author[F.~Santos]{Francisco Santos}

%\address[G.~Codenotti]
%{
%Institut f\"ur Mathematik, Freie Universit\"at Berlin, Germany
%}
%\email{codenotti@math.uni-frankfurt.de}

%\address[F.~Santos]
%{
%Department of Mathematics, Statistics and Computer Science, University of Cantabria, Spain
%}
%\email{francisco.santos@unican.es}

%\thanks{The authors were supported by the Einstein Foundation Berlin under grant EVF-2015-230. 
%Work of F. Santos is also supported by grants PID2019-106188GB-I00 funded by MCIN/AEI/10.13039/501100011033 and by FPU19/04163 of the Spanish Government and by project CLaPPo (21.SI03.64658) of Universidad de Cantabria and Banco Santander}

@@

\title{Unimodular covers of 3-dimensional parallelepipeds and Cayley sums}
\author{}

@@

@@

When the normal fan of a polytope $Q$ refines that of another polytope $P$, as in the second conjecture, we say that $P$ \emph{is a weak Minkowski summand of $Q$}, since this is easily seen to be equivalent to the existence of a polytope $P'$ such that $P+P' = k Q$ for some dilation constant $k>0$.
This property has the following algebraic implication for the projective toric variety $X_Q$: $P$ is a weak Minkowski summand of $Q$ if and only if the Cartier divisor defined by $P$ on  $X_Q$ is \emph{numerically effective}, or ``nef'' (see~\cite[Cor.~6.2.15, Prop.~6.3.12]{CLS}, but observe that what we here call  ``weak Minkowski summand'' is simply called ``Minkowski summand'' there).

@@

This has the following consequence, also proved in \Cref{sec:cayley}.
Here a  \emph{prismatoid} is a polytope whose vertices all lie in two parallel facets.

@@

Let us mention that recent work of Gubeladze~\cite{Gubeladze} shows another class of 3-polytopes admitting unimodular covers: the convex hulls of all lattice points inside an ellipsoid; these had previously been shown to be IDP by Bruns, Gubeladze and Michalek~\cite{BGM}.
To date there are no known examples of IDP $3$-polytopes without a unimodular cover, although such polytopes exist in higher dimension~\cite{BG}.

%\subsection*{Acknowledgements:} We would like to thank Akiyoshi Tsuchiya, Spencer Backman, and Johannes Hofscheier for posing these questions to us and Christian Haase for helpful discussions.

@@

Observe that $u+v=p_1+p_2=q_3+q_4$.
Now, this implies that the quadrilateral $\conv(p_1q_4p_2q_3)$ contains a fundamental domain for the lattice $\Z^2\times\{0\}$. Hence, its translate $\conv(q_2p_3q_1p_4)$ contains a fundamental domain for $\Z^2\times\{1\}$ and, in particular, it contains at least one lattice point other than $p_3$ and $p_4$. By central symmetry around its center $\left(\frac{a}2,-\frac{b}2,1\right)$, $\conv(q_2p_3q_1p_4)$ must contain lattice points in both triangles $\conv(q_2p_3p_4)\subset T_1$ and $\conv(q_1p_3p_4)\subset T_2$.

@@

The translation of vector $\frac12 (p_1+p_2-p_3-p_4)$ sends $q_1,q_2,p_3,p_4$ to $p_2,p_1,q_4,q_3$ (in this order). By applying this to inequality \eqref{eq:third}, we obtain
\[
\alpha \le f(p_2) < f(q_i) < f(p_1) \le \beta, \quad\text{ for $i=3,4$},
\]
so that $q_3,q_4 \in B$.
%
This finishes the proof of the claim, and of the lemma.

@@

\Cref{lemma:3<4} guarantees that one of the corner tetrahedra $T_i$ of the parallelepiped $C(T)$ is contained in $P$. Without loss of generality, suppose $T_4 = \conv(p_1, p_2, p_3,q_4)$ is in $P$. By \Cref{lemma:corner}, we know that $T_4$ contains a lattice point other than the $p_i$s, which we denote by $u$. 
%
Then $S=\conv(T\cup \{u\})$ can be triangulated in two different ways: $S=T \cup T'_4$, where $T'_4 = \conv(p_1, p_2, p_3, u) \subseteq T_4$ and $S= S_1 \cup S_2 \cup S_3$, with
\[
S_1= \conv(p_2,p_3,p_4, u),
S_2=\conv(p_1,p_3,p_4, u),
S_3=\conv(p_1,p_2,p_4, u).
\]

Each of the tetrahedra $S_i$ has lattice volume strictly smaller than $T$ because, for each $i$, $p_i$ is the unique point of $C(T)$ maximizing the distance to the opposite facet $\conv(p_j,p_k,p_l)$ of $T$. Thus, $S_1$, $S_2$ and $S_3$ cover $T$ and have volume strictly smaller than $T$. The $S_i$ may not be empty, but we can triangulate them into empty tetrahedra, which by inductive hypothesis they can be covered unimodularly.
\end{proof}

\begin{proof}[Proof of \Cref{thm:parallelepipeds}]
Arbitrarily triangulate the parallelepiped into empty lattice tetrahedra and apply \Cref{coro:coverpara} to these tetrahedra.
\end{proof}

Let us say that a lattice $3$-polytope $P$ \emph{has the circumscribed parallelepiped property} if it satisfies the conclusion of \Cref{lemma:3<4}:  ``for every empty tetrahedron $T$ contained in $P$ at least one of the four corner tetrahedra in $C(T)$ is 
contained in $P$''. 
If this holds then $P$ has a unimodular cover, since then the proofs of \Cref{coro:coverpara} and \Cref{thm:parallelepipeds} work for $P$.
In turn, our proof that parallelepipeds have the property (\Cref{lemma:3<4}) is based on the fact that they have only three (pairs of) normal vectors. 
In the following two examples we show a smooth $3$-polytope and two $3$-polytope with four normal vectors that do not have the property. The latter are not IDP:

@@

@@

\begin{lemma}
\label{lemma:cayley}
%\paco{$Q$ is 2-dim, but $P$ may not}
Let $Q$ be a two-dimensional lattice polytope and $P$ a weak Minkowski summand of it. 
Let $p=[p_1,p_2] \subset P$ and $q=[q_1,q_2]\subset Q$ be two primitive and non-parallel lattice segments, and let $\vecline p$ and $ \vecline q$ be the lines spanned by them.  If the parallelogram $p + q$ is not unimodular, then at least one of the regions
\[
((p_1, p_2) + \vecline q ) \cap P, 
\qquad \text{and} \qquad
((q_1, q_2) + \vecline p ) \cap Q
\]
contains a lattice point. See \Cref{fig:strips}.
\end{lemma}

@@

In turn, this means that the polytope $\conv(\tilde u, \tilde p_1, \tilde p_2, \tilde q_1, \tilde q_2) = \cayley(t,q)$
 has the following two triangulations:
\begin{gather*}
\mathcal T^+:= \left\{ \cayley(p,q), \cayley(t, \{q_i\}) \right\},
\\
\mathcal T^-:= \{ \cayley([p_1,u],q), \cayley([p_2,u],q), \cayley(t, \{q_j\}) \}.
\end{gather*}
The tetrahedra $\cayley(t, \{q_j\})$ and $\cayley(t, \{q_i\})$ are unimodular, which implies that $T=\cayley(p,q)$ has volume equal to the sum of the volumes of $\cayley([p_1,u],q)$ and $\cayley([p_2,u],q)$. In particular, we have covered $T$ by the three tetrahedra in $\mathcal T^-$, which are of smaller volume and hence have unimodular covers by inductive assumption.
\end{proof}

@@

\begin{proof}[Proof of \Cref{coro:prismatoid}]
The polytope under study satisfies the hypotheses of \Cref{prop:prismatoid}: %in \Cref{coro:prismatoid}, 
the smoothness of the prismatoid implies that every edge of the form $[u\times \{0\}, v\times \{k\}]$ has lattice points in all slices. 
Hence,
\[
kP \cap(\R^2\times\{1\}) = (k-1)Q_1 + Q_2.
\qedhere
\] 
\end{proof}

@@

@@

@@

@@

\begin{figure}[htb]
\scalebox{.75}{\input{claim3.pdf_t}}
\caption{Illustration of the proof of \Cref{claim:r}}
\label{fig:claim3}
\end{figure}

@@

@@

\begin{claim}
\label{claim:g}
$g(t_1) > g(t_2)$, $g(b_1) > g(b_2)$, and $g(r_1) < g(r_2)$.
\end{claim}

@@
